---
title: "VDAC1 in Alzheimer's Disease: A Cross-Dataset Transcriptomic Meta-Analysis"
author: "Jeyakumaran Shanmugam (JK)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(knitr)
library(metafor)
library(ggplot2)
```

## Introduction

Voltage-dependent anion channel 1 (VDAC1) is a mitochondrial outer membrane protein involved in metabolite exchange, apoptosis, and cellular energy regulation. Mitochondrial dysfunction is a hallmark of Alzheimer's disease (AD). I performed a systematic meta-analysis of publicly available human brain transcriptomic datasets to determine whether VDAC1 expression is altered in AD and if changes relate to sex or neuropathological severity.

## Methods

I identified eight independent post-mortem human brain transcriptomic datasets from NCBI GEO with available raw/normalized expression data and sample metadata. VDAC1 expression was extracted using platform-specific probe mappings. Expression values were log2-transformed. For each dataset, I computed the log2 fold change (AD vs. control) and standard error. Random-effects meta-analysis was performed using the `metafor` R package. Sex-stratified analyses were conducted separately. Spearman correlation assessed the relationship between VDAC1 expression and Braak neuropathological stage.

## Results

### Primary Meta-Analysis (All Samples)

```{r primary}
overall <- read.csv("results/overall_meta_summary.csv")
kable(overall, digits = 3, caption = "Overall random-effects meta-analysis")
```

```{r forest_overall, fig.width=10, fig.height=6}
knitr::include_graphics("results/forest_overall.pdf")
```

### Sex-Stratified Meta-Analysis

```{r sex}
sex_sum <- read.csv("results/sex_meta_summary.csv")
kable(sex_sum, digits = 3, caption = "Meta-analysis by sex")
```

**Female**:

```{r forest_female, fig.width=10, fig.height=5}
knitr::include_graphics("results/forest_female.pdf")
```

**Male**:

```{r forest_male, fig.width=10, fig.height=5}
knitr::include_graphics("results/forest_male.pdf")
```

### Neuropathological Severity (Braak Stage)

```{r braak_corr}
braak_corr <- read.csv("results/braak_correlation_results.csv")
kable(braak_corr, digits = 3, caption = "Spearman correlation between VDAC1 and Braak stage")
```

```{r braak_boxplot, fig.width=10, fig.height=6}
knitr::include_graphics("results/braak_boxplot.pdf")
```

## Conclusion

VDAC1 expression is significantly downregulated in AD brains across multiple independent cohorts, with evidence for a stronger effect in females. Moreover, VDAC1 levels negatively correlate with Braak neuropathological stage, suggesting a possible link between mitochondrial dysfunction and disease progression. These findings support further investigation of VDAC1 as a potential biomarker or therapeutic target in Alzheimer's disease.

## Reproducibility

All data were obtained from public GEO repositories. Analysis scripts and processed data are available in the accompanying GitHub repository.
